### 2018-12-13 chapters/1-introduction.tex
citet -> cite

Computer-aided diagnostics (CAD) based on MRI has been demonstrated to have
complementary role to a reporting radiologist in PCa detection \citep{Kwak2015,
Viswanath2012, Ginsburg2014}. However, only a limited number of studies focused
on characterizing the detected PCa lesions, and they typically utilize only a
small number of texture features. \citet{Tiwari2013} used MR spectroscopy data
and various texture features from T₂w to first detect PCa and then predict its
Gleason score. \citet{Peng2013} evaluated histogram-based features from
multiparametric MRI regarding correlation with Gleason score. Texture features
from DWI and T₂w have been assessed for differentiating Gleason scores
\citep{Wibmer2015, Vignati2015, Fehr2015}. \citet{Rozenberg2016} evaluated
whole-lesion histogram and texture features from DWI in order to predict Gleason
score upgrade after radical prostatectomy.

>>>>

Computer-aided diagnostics (CAD) based on MRI has been demonstrated to have
complementary role to a reporting radiologist in PCa detection \citep{Kwak2015,
Viswanath2012, Ginsburg2014}. However, only a limited number of studies focused
on characterizing the detected PCa lesions, and they typically utilize only a
small number of texture features. Tiwari et al.\ \cite{Tiwari2013} used MR
spectroscopy data and various texture features from T₂w to first detect PCa and
then predict its Gleason score. Peng et al.\ \cite{Peng2013} evaluated
histogram-based features from multiparametric MRI regarding correlation with
Gleason score. Texture features from DWI and T₂w have been assessed for
differentiating Gleason scores \citep{Wibmer2015, Vignati2015, Fehr2015}.
Rozenberg et al.\ \cite{Rozenberg2016} evaluated whole-lesion histogram and
texture features from DWI in order to predict Gleason score upgrade after
radical prostatectomy.

### 2018-12-13 chapters/2-methods.tex
table grooming
citet -> cite

The signal intensity of a T₂w image is not a specific tissue property and
this non-standardness of T₂w images (``intensity drift'') requires
standardization to a common scale. To correct this bias, a histogram alignment
method was used, as described by \citet{Nyul1999,Nyul2000}. This simple method
transforms the images to make their histograms match at certain landmark
locations,  and interpolates the values between.

In this study, the deciles (i.e.\ every tenth percentile) were used as
landmarks, as suggested by \citet{Nyul1999}. Only delineated prostate volume was
considered for histogram averages, as other parts of the image might distort the
learning. The result was validated by visual inspection.

@@

Several texture descriptor methods with various parameter combinations were used
for extracting 2D texture features from the manually delineated PCa lesions.
Most of the methods in this study inherently incorporate the so-called ``sliding
window'' algorithm. This means that the local voxel neighborhood for
calculations is represented as a fixed-shape subwindow centered at each voxel in
turn. The output image, a texture feature map, consists of the feature values
positioned on corresponding neighborhood center locations. A more detailed
explanation is provided by \citet{Clausi2002Rapid}, for example.

@@

The gray-level co-occurrence matrix (GLCM) \citep{Haralick1973} is a very
popular method of texture characterization. It observes all the pixel gray level
pairings that occur in the image at a certain distance and direction. In this
study, the GLCM was calculated for each window using four different voxel
distances of 1 to 4 (unless prevented by window dimensions). Because the 22
GLCM-derived features introduced by \citet{Haralick1973} have correlation
\citep{Albregtsen2008}, using only three to five features have been recommended
\citep{Clausi2002Analysis, Albregtsen2008, Gebejes2013a} in order to minimize
redundancy and dimensionality.

@@

In addition to the sliding window approach, the same GLCM-based features were
also extracted using the minimum bounding box (MBB) around the whole lesion
area of each image slice as the window, while ignoring any non-lesion voxels.
This procedure is similar to the method used in a study by \citet{Vignati2015}.
The MBB-GLCM features were averaged over slices, and 48 features in total were
extracted per image type.

@@

The Gabor function \citep{Gabor1946} is a Gaussian modulated by a sinusoidal
wave. Gabor filter banks can be used for texture characterization, as each
filter, shaped by its parameters, responds to specific local spatial frequency
properties of the image \citep{Turner1986}. Gabor filters are sensitive to edges
in the image, so given that different lesions contain different regions, the
detected edges between the regions could yield different responses. In this
study, the texture extraction scheme described by \citet{Tuceryan1998} was
applied, which uses the sliding window on the Gabor-filtered complex images.

@@

Afterwards, each data point was assigned an LPOCV score according to the
ordering-by-the-number-of-wins method, in which the score of a data point is the
number of times it obtains a larger predicted value than the other point during
the LPOCV rounds when it is one of the two held out points. These LPOCV scores
were then used to perform the ROC curve analysis, whose validity was previously
demonstrated by \citet{Balcan2008}. More precisely, using the LPOCV scores of
the data points and their corresponding true label, the AUC and 95\% confidence
interval (CI) were calculated using the R package by \citet{LeDell2015}.

@@

\begin{algorithm}[!h]
  \caption{LPOCV with inner feature selection by AUC filtering and
  hyperparameter selection.}%
  \label{alg:pseudocode}

>>>>

The signal intensity of a T₂w image is not a specific tissue property and
this non-standardness of T₂w images (``intensity drift'') requires
standardization to a common scale. To correct this bias, a histogram alignment
method was used, as described by Nyúl et al.\ \cite{Nyul1999, Nyul2000}. This
simple method transforms the images to make their histograms match at certain
landmark locations, and interpolates the values between.

In this study, the deciles (i.e.\ every tenth percentile) were used as
landmarks, as suggested by Nyúl et al.\ \cite{Nyul1999}. Only delineated
prostate volume was considered for histogram averages, as other parts of the
image might distort the learning. The result was validated by visual inspection.

@@

Several texture descriptor methods with various parameter combinations were used
for extracting 2D texture features from the manually delineated PCa lesions.
Most of the methods in this study inherently incorporate the so-called ``sliding
window'' algorithm. This means that the local voxel neighborhood for
calculations is represented as a fixed-shape subwindow centered at each voxel in
turn. The output image, a texture feature map, consists of the feature values
positioned on corresponding neighborhood center locations. A more detailed
explanation is provided by Clausi et al.\ \cite{Clausi2002Rapid}, for example.

@@

The gray-level co-occurrence matrix (GLCM) \citep{Haralick1973} is a very
popular method of texture characterization. It observes all the pixel gray level
pairings that occur in the image at a certain distance and direction. In this
study, the GLCM was calculated for each window using four different voxel
distances of 1 to 4 (unless prevented by window dimensions). Because the 22
GLCM-derived features introduced by Haralick et al.\ \cite{Haralick1973} have
correlation \citep{Albregtsen2008}, using only three to five features have been
recommended \citep{Clausi2002Analysis, Albregtsen2008, Gebejes2013a} in order to
minimize redundancy and dimensionality.

@@

In addition to the sliding window approach, the same GLCM-based features were
also extracted using the minimum bounding box (MBB) around the whole lesion
area of each image slice as the window, while ignoring any non-lesion voxels.
This procedure is similar to the method used in a study by Vignati et
al.\ \cite{Vignati2015}. The MBB-GLCM features were averaged over slices, and
48 features in total were extracted per image type.

@@

The Gabor function \citep{Gabor1946} is a Gaussian modulated by a sinusoidal
wave. Gabor filter banks can be used for texture characterization, as each
filter, shaped by its parameters, responds to specific local spatial frequency
properties of the image \citep{Turner1986}. Gabor filters are sensitive to edges
in the image, so given that different lesions contain different regions, the
detected edges between the regions could yield different responses. In this
study, the texture extraction scheme described by Tüceryan et
al.\ \cite{Tuceryan1998} was applied, which uses the sliding window on the
Gabor-filtered complex images.

@@

Afterwards, each data point was assigned an LPOCV score according to the
ordering-by-the-number-of-wins method, in which the score of a data point is the
number of times it obtains a larger predicted value than the other point during
the LPOCV rounds when it is one of the two held out points. These LPOCV scores
were then used to perform the ROC curve analysis, whose validity was previously
demonstrated by Balcan et al.\ \cite{Balcan2008}. More precisely, using the
LPOCV scores of the data points and their corresponding true label, the AUC and
95\% confidence interval (CI) were calculated using the R package by LeDell et
al.\ \cite{LeDell2015}.

@@

\begin{algorithm}[!h]
  \caption{{\bf LPOCV with inner feature selection by AUC filtering and
  hyperparameter selection.}}%
  \label{alg:pseudocode}

### 2018-12-13 chapters/4-discussion.tex
citet -> cite

There is an increasing number of research groups studying and developing CAD of
prostate cancer. The topic was recently reviewed by \citet{Lemaitre2015}. Most
of the publications have focused on the task of cancer detection rather than
characterization. However, both are required for proper treatment decision
planning. In this study, we built a classifier for prostate cancer
characterization utilizing texture features extracted from T₂w, the
monoexponential and kurtosis models of high-b-value DWI, and T₂ maps.

@@

Texture analysis of multiparametric MRI has previously seen limited use in PCa
characterization. \citet{Fehr2015} evaluated PCa characterization with the
whole-lesion first order statistics and GLCM texture features from a
similar-size dataset of ADC and T₂w. They used oversampling to ward off effect
of class imbalance. Similarly to our study, they integrated dynamic feature
selection as part of the training (using the recursive feature selection support
vector machine, RFE-SVM). In our study, we included a much more diverse and
numerous set of features, as one of our goals was to evaluate various texture
extraction methods. Moreover, we have for the first time demonstrated that using
texture features from K (kurtosis function) provided improvements to ADCₘ
(monoexponential function). This is important since first order statistics of
parameters derived from kurtosis function do not lead to improved performance of
ADCm (monoexponential function). The effect of noise remains to be explored in
future studies.

\citet{Tiwari2013} classified PCa using GLCM and simple gradient features from
T₂w and MR spectroscopy (MRS). A multi-kernel classifier with graph embedding
was used to reduce dimensionality. Compared to the current study, they had fewer
patients, and the classification was done on equally-sized, rectangular
metavoxels.

Furthermore, \citet{Wibmer2015} studied the associations of Gleason scores and
individual GLCM features from ADC and T₂w of PZ lesions, using generalized
linear regression and generalized estimating equations; and \citet{Vignati2015}
tested Gleason score differentiation using two of the GLCM features (contrast
and homogeneity) from T₂w and ADC individually.

@@

We highlight the limitation of performing re-slicing and
non-rigid deformation of MR data sets to common space and then co-registering
with whole mount prostatectomy sections. As correctly noted by
\citet{Bourne2017}, co-registration of whole mount prostatectomy sections to MRI
data sets is important. However, the effect of re-slicing and non-rigid
deformation of MR data sets to common space remains to be explored.

@@

The image moments, on the other hand, have been used more often for global
morphological analysis like shape recognition rather than local texture
analysis, although they have been used for texture as well~\citep{Tuceryan1994,
Laws1980, Tuceryan1990}. To the best of our knowledge, there are no published
studies using moment-based texture analysis for detection or characterization of
prostate cancer using MRI data sets. \citet{Tahmasbi2011} used Zernike moments
to characterize breast cancer, but as a global mass descriptor and not for
texture. Our results suggest that moment-based texture features might be
valuable for PCa characterization. More specifically, the best 1\% features of
the image types K and T₂, and the final model ADCₘ, K, and T₂w combined
included some of the texture features based on Hu or Zernike moments.

>>>>

There is an increasing number of research groups studying and developing CAD of
prostate cancer. The topic was recently reviewed by Lemaître et
al.\ \cite{Lemaitre2015}. Most of the publications have focused on the task of
cancer detection rather than characterization. However, both are required for
proper treatment decision planning. In this study, we built a classifier for
prostate cancer characterization utilizing texture features extracted from T₂w,
the monoexponential and kurtosis models of high-b-value DWI, and T₂ maps.

@@

Texture analysis of multiparametric MRI has previously seen limited use in PCa
characterization. Fehr et al.\ \cite{Fehr2015} evaluated PCa characterization
with the whole-lesion first order statistics and GLCM texture features from a
similar-size dataset of ADC and T₂w. They used oversampling to ward off effect
of class imbalance. Similarly to our study, they integrated dynamic feature
selection as part of the training (using the recursive feature selection support
vector machine, RFE-SVM). In our study, we included a much more diverse and
numerous set of features, as one of our goals was to evaluate various texture
extraction methods. Moreover, we have for the first time demonstrated that using
texture features from K (kurtosis function) provided improvements to ADCₘ
(monoexponential function). This is important since first order statistics of
parameters derived from kurtosis function do not lead to improved performance of
ADCm (monoexponential function). The effect of noise remains to be explored in
future studies.

Tiwari et al.\ \cite{Tiwari2013} classified PCa using GLCM and simple gradient
features from T₂w and MR spectroscopy (MRS). A multi-kernel classifier with
graph embedding was used to reduce dimensionality. Compared to the current
study, they had fewer patients, and the classification was done on
equally-sized, rectangular metavoxels.

Furthermore, Wibmer et al.\ \cite{Wibmer2015} studied the associations of
Gleason scores and individual GLCM features from ADC and T₂w of PZ lesions,
using generalized linear regression and generalized estimating equations; and
Vignati et al.\ \cite{Vignati2015} tested Gleason score differentiation using
two of the GLCM features (contrast and homogeneity) from T₂w and ADC
individually.

@@

We highlight the limitation of performing re-slicing and
non-rigid deformation of MR data sets to common space and then co-registering
with whole mount prostatectomy sections. As noted by Bourne et
al.\ \cite{Bourne2017}, co-registration of whole mount prostatectomy sections to
MRI data sets is important. However, the effect of re-slicing and non-rigid
deformation of MR data sets to common space remains to be explored.

@@

The image moments, on the other hand, have been used more often for global
morphological analysis like shape recognition rather than local texture
analysis, although they have been used for texture as well~\citep{Tuceryan1994,
Laws1980, Tuceryan1990}. To the best of our knowledge, there are no published
studies using moment-based texture analysis for detection or characterization of
prostate cancer using MRI data sets. Tahmasbi et al.\ \cite{Tahmasbi2011} used
Zernike moments to characterize breast cancer, but as a global mass descriptor
and not for texture. Our results suggest that moment-based texture features
might be valuable for PCa characterization. More specifically, the best 1\%
features of the image types K and T₂, and the final model ADCₘ, K, and T₂w
combined included some of the texture features based on Hu or Zernike moments.
